A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-11-12 model.tex
notation change for dg vel diff
home/work sync

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\Delta\bm{v}}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume, where
\begin{align}
  \Delta\bm{v} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
\end{align}
is the dust-gas velocity difference; and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density. 
Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rhog/\rho$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

In general $\Delta\bm{v}$ obeys a complicated evolutionary equation
\citep[see, e.g.][]{youdin05a}. However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \Delta\bm{v} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{youdin05a,jacquet11}. This equation implies that dust particles
drift towards pressure maxima.

@@

In Eq. \ref{dusteq}, 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. Note that Eq. \ref{dusteq} is 
equivalent to Eq. 46 of \cite{jacquet11}.

@@

The gas component of the mixture has a specific entropy 
$S = C_p\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_p$ is the heat 
capacity at constant pressure. However, this is not a useful quantity
for (locally) isothermal gas.

Nevertheless, since we have shown that an isothermal dusty gas
effectively has $\gamma=1$ we can define an effective (dimensionless)  
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then 
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is just the statement of entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag instabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/}}.

@@

\subsubsection{Physical picture of dust-gas drag instabilities}

>>>>

The two fluids interact via a drag force parameterized by a stopping
time $\tstop$ 
such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\widetilde{\bm{v}}}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume, where
\begin{align}
  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
\end{align}
is the dust-gas velocity difference; and 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density. 
Note that $\tstop$ 
differs slightly from the stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a,jacquet11}.

In general $\widetilde{\bm{v}}$ obeys a complicated evolutionary equation
\citep[see, e.g.][]{youdin05a}. However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \widetilde{\bm{v}} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation implies that dust particles
drift towards pressure maxima.

@@

Eq. \ref{dusteq} results from mass conservation of dust, where 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. Note that Eq. \ref{dusteq} is 
equivalent to Eq. 46 of \cite{jacquet11}.

@@

The specific entropy of an ideal gas is 
$S = C_p\ln{\left(P^{1/\gamma}/\rhog\right)}$, where $C_p$ is the heat 
capacity at constant pressure.

Since we have shown that a locally isothermal dusty gas effectively has $\gamma=1$ we can define an effective (dimensionless)   
entropy for the \emph{mixture} as 
\begin{align}
   \seff \equiv \ln \frac{P}{\rho} = \ln{\left[c_s^2(1-\tepsilon)\right]}.  
\end{align} 
%This follows naturally by combining the effective 
%energy Eq. \ref{eff_energy} and Eq. \ref{masseq}. 
Then 
\begin{align*}
  \frac{D \seff}{D t} = \bm{v}\cdot\nabla\ln{c_s^2} +
  \frac{c_s^2}{P}\nabla\cdot\left(\tstop\tepsilon\nabla P\right), 
\end{align*}
which is just the statement of entropy evolution in standard  
hydrodynamics, albeit with source terms. With an effective
entropy defined this way, many of the  
results concerning the stability of (locally) isothermal dusty gas
will have identical form and interpretations to that for pure gas.

@@

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag instabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/print-chap-full.pdf}}.

@@

\subsubsection{Physical requirement of dust-gas drag instabilities}
